---
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{IntakeBalance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


`IntakeBalance` is an R package to help with implementing intake-balance methods
for assessing energy intake (EI). In this vignette, we will demonstrate how to
do this using an accelerometer-based approach. The goal is to make these tools
and methods as accessible as possible, and we understand that most potential
users are not R programmers. A certain level of interaction with R will be
required to implement the methods, but we aim to provide thorough instructions
that will demystify the process and limit how intimidating it feels. On that
note, revisions may be necessary to smooth out areas that are not as clear as we
initially hoped. Please feel free to point these areas out to us by
[filing an issue](https://github.com/paulhibbing/IntakeBalance/issues/new/choose).


## Prerequisites

Before you can use `IntakeBalance`, you need to get your machine set up. The
first step is to [install R](https://cran.r-project.org/). We also recommend
[installing RStudio](https://posit.co/download/rstudio-desktop/). The default
settings should do the trick, so you can just run the installers. Then open up
RStudio and prepare to run some code in the console. (The bottom left pane. Code
can be pasted in like it would in a document. Then press `[enter]` to execute
the code.)

The next task is to install the `IntakeBalance` package. This should only be one
or two commands, but we're going to go the long way in hopes of avoiding any
breakdowns that might depend on your specific computer setup or prior
installation of R. Just copy, paste, and execute all of the following code:

```{r, installation, eval=FALSE}

## List the required packages to install from mainstream R archive
cran_packages <- c(
  "digest", "dplyr", "knitr", "lubridate", "magrittr",
  "nnet", "PAutilities", "PhysActBedRest", "PhysicalActivity", "purrr", "randomForest",
  "read.gt3x", "devtools", "rmarkdown", "rlang",
  "Sojourn", "stats", "testthat", "tidyr", "tree",
  "TwoRegression", "units", "utils", "xml2"
)


## Install any of the required packages that aren't already installed
invisible(lapply(
  cran_packages,
  function(x) if (!isTRUE(requireNamespace(x, quietly = TRUE)))
    install.packages(x)
))


## List the required packages to install from GitHub
remote_packages <- c(
  "paulhibbing/anthropometry", "paulhibbing/accelEE",
  "paulhibbing/EE.Data", "paulhibbing/IntakeBalance"
)


## Install any of the GitHub packages that aren't already installed
invisible(lapply(
  remote_packages,
  function(x) if (!isTRUE(requireNamespace(x, quietly = TRUE)))
    devtools::install_github(x, dependencies = FALSE)
))


## Install one other package using a different method,
## if it hasn't already been installed

if (!requireNamespace("AGread", quietly = TRUE))
  devtools::install_version("AGread", "1.3.2")

```

Once you've finished the above, you're ready to look at an example of using the
`IntakeBalance` package.


## Basic Usage

The core of `IntakeBalance` is an eponymous function that simply calculates EI
based on information about body composition and energy expenditure. The function
runs in two ways depending on what information you supply. You can read the
documentation about the function by running `IntakeBalance::IntakeBalance`.
Let's look at some imaginary data to show how this works.

#### 1. The vector method

This is designed for making calculations when you're direcly passing in data
values for the requested pieces of information (i.e., "parameters" or
"arguments"). It looks like this:

```{r, vector}
IntakeBalance::IntakeBalance(
  fm_start = 50.5,   ## Fat mass at the start of the measurement period
  fm_end = 50.7,     ## Fat mass at the end of the measurement period
  ffm_start = 75.1,  ## Fat-free mass at the start of the measurement period
  ffm_end = 74.9,    ## Fat-free mass at the end of the measurement period
  fm_units = "kg",   ## Units of measure for the fat mass value
  ffm_units = "kg",  ## Units of measure for the fat-free mass value
  ee_per_day = 1950, ## Daily energy expenditure
  ee_units = "kcal", ## Units of measure for the resting energy expenditure value
  n_days = 14        ## Number of days between the start and end measurements
)
```
This tells us that calculated energy intake was 2071 kcal/day.

#### 2. The data frame method

This is designed for making calculations when you have data for multiple
individuals stored in a data frame. To show how it works, we first need to
create some imaginary data:

```{r, df}
## Create the data frame
info <- data.frame(
  fm_start = c(50.5, 70.2), fm_end = c(50.7, 70.0),
  ffm_start = c(75.1, 90.3), ffm_end = c(74.9, 90.3),
  ee = c(1950, 2473), n_days = c(14, 14)
)

## Show the data frame
print(info)
```

Now let's calculate EI:

```{r, df2}
IntakeBalance::IntakeBalance(
  
  ## These arguments still refer to the same information outlined previously,
  ## but now they refer to the names of columns containing the values, rather
  ## than providing the values themselves
  fm_start = "fm_start",
  fm_end = "fm_end",
  ffm_start = "ffm_start",
  ffm_end = "ffm_end",
  ee_per_day = "ee",
  n_days = "n_days",
  
  ## The trick is to pass in a data frame via this extra argument. That's how R
  ## knows to interpret the other variables as column names rather than
  ## expecting values to be passed in directly.
  df = info
  
)
```

You can see that this returns the original data frame, updated with new columns
that indicate the change in energy storage (`delta_ES`) and the EI values.
Furthermore, you can see that the values in the first row of `info` match with
what was provided in the earlier example.


## Making this Work with Data from Wearables

The above examples demonstrated the core functionality of the `IntakeBalance`
package. But things get more powerful when we combine this with other packages
to help with wrangling wearable data and incorporating it into the
intake-balance method. Here, we will show how you could use the `accelEE`
package to process an ActiGraph file using the same methods from our doubly
labeled water study.

The first step is to read a "raw data" file (`gt3x` extension) into R. We'll use
one that was built in with one of the packages we installed at the beginning. It
only contains 40 min of data, so we'll copy it several times and pretend it
spans several days instead.

```{r, example}

## Find the file
filename <- system.file(
  "extdata/TAS1H30182785_2019-09-17.gt3x",
  package = "read.gt3x"
)

## Read the file
ex_data <- accelEE::ee_file(filename = filename, scheme = "Hibbing 2023")

## Extend the data file (ignore these commands)
ex_data <- data.frame(
  Timestamp = ex_data$Timestamp[1] + (60*0:3199),
  do.call(rbind, replicate(80, ex_data[ ,-1], simplify = FALSE))
)

## Show some of the contents
head(dplyr::select(ex_data, !matches("features")))

```

So far, this gives us minute-by-minute summaries of accelerometer variables,
plus estimates of energy expenditure for the different methods. Next, we need to
account for non-wear and sleep. If you have a file with `agd` extension, you can
use syntax just like the above to acquire this information. For example, you
might run `agd_data <- accelEE::ee_file(filename = "myfolder/myfile.agd", scheme
= "Hibbing 2023")`. R will recognize that it is an `agd` file and not a `gt3x`
file, and it will then run the desired non-wear and sleep procedures instead of
generating estimates of energy expenditure.

In this case, we don't have a built in `agd` file, so we'll
just make up some activity count data like this:

```{r, example2}

## Set a starting point for the random number generator so that sampling is
## repeatable when generating the `Axis1` and `valid_status` columns
set.seed(610)

## Now generate the sleep/non-wear dataframe
nonwear_data <- data.frame(
  Timestamp = ex_data$Timestamp,
  Axis1 = sample(0:200, 3200, TRUE, exp(rnorm(201, log(30), log(5)))),
  Axis2 = 0,
  Axis3 = 0,
  valid_status = sample(
    c("Awake-Wear", "Non-Wear", "Sleep"),
    3200,
    TRUE,
    c(2, 8, 14)
  )
)
```

Then we can merge the two together and get a daily summary of EE:

```{r, example3}

full_data <- merge(ex_data, nonwear_data)

accelEE::ee_summary(full_data, scheme = "Hibbing 2023")

```

This gives us estimates of daily energy expenditure, in kcal/kg. Now we need to
combine it with anthropometric data so we can determine change in energy storage
and ultimately calculate EI.

We need to account for non-wear and sleep data by imputing basal metabolic rate estimates.
