---
title: Using the `IntakeBalance` Package
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{IntakeBalance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


`IntakeBalance` is an R package to help with implementing intake-balance methods
for assessing energy intake (EI). In this vignette, we will demonstrate how to
do this using an accelerometer-based approach.

We realize this might seem like an intimidating tool, since most potential users
are not R programmers. Although it's true that we can currently only provide
support for using this tool in R, we want to make it as accessible as possible.
On this page, our goal is to provide thorough instructions that will demystify
the process and limit how intimidating it feels. But we also recognize there may
be areas where things need to be clearer. If you find anything confusing, please
feel free to
[let us know by filing an issue](https://github.com/paulhibbing/IntakeBalance/issues/new/choose),
and we will do our best to revise in a way that makes these instructions easier
to follow.


## Prerequisites

Before you can use `IntakeBalance`, you need to get your machine set up. The
first step is to [install R](https://cran.r-project.org/) if you haven't
already. We also recommend [installing
RStudio](https://posit.co/download/rstudio-desktop/). The default settings
should do the trick, so you can just run the installers. Then open up RStudio
and prepare to run some code in the console. (The bottom left pane. Code can be
pasted in like it would in a document. Then press `[enter]` to execute the
code.)

The next task is to install the `IntakeBalance` package. This could be done much
more concisely than what's shown in the code below, but we're going to go the
long way in hopes of avoiding any breakdowns that might occur. (Sometimes things
work differently depending on your specific computer setup or prior installation
of R, so this longer approach helps us to [hopefully] account for that.) Just
copy, paste, and execute all of the following code:

```{r, installation, eval=FALSE}
## List the required packages to install from mainstream R archive
cran_packages <- c(
  "digest", "dplyr", "knitr", "lubridate", "magrittr",
  "nnet", "PAutilities", "PhysActBedRest", "PhysicalActivity",
  "purrr", "randomForest", "read.gt3x", "devtools", "rmarkdown",
  "rlang", "Sojourn", "stats", "testthat", "tidyr", "tree",
  "TwoRegression", "units", "utils", "xml2"
)


## Install any of the previous packages that aren't already installed
invisible(lapply(
  cran_packages,
  function(x) if (!isTRUE(requireNamespace(x, quietly = TRUE)))
    install.packages(x)
))


## List the required packages to install from GitHub
remote_packages <- c(
  "paulhibbing/anthropometry", "paulhibbing/accelEE",
  "paulhibbing/EE.Data", "paulhibbing/IntakeBalance"
)


## Install any of the GitHub packages that aren't already installed
invisible(lapply(
  remote_packages,
  function(x) if (!isTRUE(requireNamespace(x, quietly = TRUE)))
    devtools::install_github(x, dependencies = FALSE)
))


## Install one other package using a different method,
## if it hasn't already been installed

if (!requireNamespace("AGread", quietly = TRUE))
  devtools::install_version("AGread", "1.3.2")
```

Once you've finished the above, you're ready to look at an example of using the
`IntakeBalance` package.


## Basic Usage

The core of `IntakeBalance` is an eponymous function that simply calculates EI
based on information about energy expenditure and changes in body composition.
As shown below, the function runs in two ways depending on what information you
supply. You can read more about this in the function documentation -- Just run
<br /> `?IntakeBalance::IntakeBalance`.

Let's look at some imaginary data to show both ways this can work.

#### Option 1. The vector method

This is designed for making calculations when you're directly passing in data
values for the requested pieces of information (i.e., "parameters" or
"arguments"). It looks like this:

```{r, vector}
## Generate the result
vector_result <- IntakeBalance::IntakeBalance(
  fm_start = 50.5,   ## Fat mass at the start of the measurement period
  fm_end = 50.7,     ## Fat mass at the end of the measurement period
  ffm_start = 75.1,  ## Fat-free mass at the start of the measurement period
  ffm_end = 74.9,    ## Fat-free mass at the end of the measurement period
  fm_units = "kg",   ## Units of measure for the fat mass value
  ffm_units = "kg",  ## Units of measure for the fat-free mass value
  ee_per_day = 1950, ## Daily energy expenditure
  ee_units = "kcal", ## Units of measure for the daily energy expenditure value
  n_days = 14        ## Number of days between the start and end measurements
)

## Show the result
knitr::kable(vector_result)
```

This tells us that calculated EI was 2071 kcal/day.

#### Option 2. The data frame method

This is designed for making calculations when you have data for multiple
individuals stored in a data frame. To show how it works, we first need to
create some imaginary data:

```{r, df}
## Create the data frame
info <- data.frame(
  PID = c("001", "002"),
  fm_start = c(50.5, 70.2), fm_end = c(50.7, 70.0),
  ffm_start = c(75.1, 90.3), ffm_end = c(74.9, 90.3),
  ee = c(1950, 2473), n_days = c(14, 14)
)

## Show the data frame
knitr::kable(info)
```

Now let's calculate EI:

```{r, df2}
## Generate the result
df_result <- IntakeBalance::IntakeBalance(
  
  ## These arguments still refer to the same information outlined previously,
  ## but now we have added a layer of abstraction to reference the names of
  ## columns where that information is stored, rather than providing the
  ## values themselves
  fm_start = "fm_start",
  fm_end = "fm_end",
  ffm_start = "ffm_start",
  ffm_end = "ffm_end",
  ee_per_day = "ee",
  n_days = "n_days",
  
  ## The trick is to pass in a data frame via this extra argument. That's how R
  ## knows to interpret the other variables as column names rather than raw values
  df = info
  
)

## Show the result
knitr::kable(df_result)
```

You can see that this returns the original data frame, updated with new columns
that indicate the change in energy storage (`delta_ES`) and the EI values.
Furthermore, you can see that the values in the first row of `info` match with
what was provided in the example for Option 1. Cool!


## Making this Work with Data from Wearables

The above examples demonstrated the core functionality of the `IntakeBalance`
package. But things get more powerful when we combine this with other packages
to help with wrangling wearable data and incorporating it into the
intake-balance framework. Here, we will show how you could use the `accelEE`
package to process an ActiGraph file using the same methods from our doubly
labeled water study. Unfortunately, the code is designed for combining `gt3x`
and `agd` files (raw acceleration and activity count formats, respectively) that
each contain at least one full day of data. This is an issue because sample
files tend to be small and only available in one format. So, throughout the code
below, you will see some extra steps we need to take, just to create a large,
properly-formatted, illustration-ready dataset that plays nicely with the code.
Don't let this throw you off. We'll tell you whenever this these extraneous
steps are happening, and you won't need to worry about them when dealing with
your own data from the real world.

The accelerometer-based intake-balance process unfolds in three phases outlined
below.

#### Phase 1. Reading and pre-formatting the wearable data

The first step is to read a "raw data" file (`gt3x` extension) into R. We'll use
one that was built in with one of the packages we installed at the beginning. It
only contains 40 min of data, so we'll copy it several times and pretend it
spans several days instead.

```{r, example}

## Find the file
filename <- system.file(
  "extdata/TAS1H30182785_2019-09-17.gt3x",
  package = "read.gt3x"
)

## Read the file using the `accelEE` package
ex_data <- accelEE::ee_file(filename = filename, scheme = "Hibbing 2023")

## Extend the data file for the sake of illustration
## (ignore this part when processing real-world data)
ex_data <- data.frame(
  Timestamp = ex_data$Timestamp[1] + (60*0:3199),
  do.call(rbind, replicate(80, ex_data[ ,-1], simplify = FALSE))
)

## Show some of the contents
knitr::kable(head(dplyr::select(ex_data, !matches("features"))))

```

So far, this gives us minute-by-minute summaries of accelerometer variables,
plus estimates of energy expenditure (kcal/kg) for the different methods. Next,
we need to account for non-wear and sleep. If you have a file with `agd`
extension, you can use syntax just like the above to acquire that information.
For example, you might run `agd_data <- accelEE::ee_file(filename =
"myfolder/myfile.agd", scheme = "Hibbing 2023")`. R will recognize that it is an
`agd` file and not a `gt3x` file, and it will then run the desired non-wear and
sleep procedures instead of generating estimates of energy expenditure.

For this example, though, we don't have a built in `agd` file. Instead, we'll
just make up some activity count data. (You're welcome to run the below code if
you're following along with this example, but in an actual application, as
described previously, you would use an `agd` file rather than completing this
step.)

```{r, example2}

## Set a starting point for the random number generator so that sampling is
## repeatable when generating the `Axis1` and `valid_status` columns
set.seed(610)

## Now generate the sleep/non-wear dataframe
agd_data <- data.frame(
  Timestamp = ex_data$Timestamp,
  Axis1 = sample(0:200, 3200, TRUE, exp(rnorm(201, log(30), log(5)))),
  Axis2 = 0,
  Axis3 = 0,
  valid_status = sample(
    c("Awake-Wear", "Non-Wear", "Sleep"),
    3200,
    TRUE,
    c(2, 8, 14)
  )
)
```

Once we have data related to both energy expenditure (`.gt3x` origin) and
sleep/non-wear (`.agd` origin), we can merge the two together and get a daily
summary:

```{r, example3}

full_data <- merge(ex_data, agd_data)

ee_data <- accelEE::ee_summary(full_data, scheme = "Hibbing 2023")

knitr::kable(ee_data)

```

#### Phase 2. Getting total daily energy expenditure from the wearable data

The previous step gave us daily estimates of total energy expenditure, expressed
relative to body mass (kcal/kg) and exclusively reflecting measurement during
awake wear time. To calculate total (absolute) energy expenditure for the full
day (including sleep/non-wear), we need to bring in our body composition data
and make another calculation or two.

```{r, merge}

## First, let's get our body composition data. We'll pretend it's the first row
## of the `info` data frame we created earlier. We'll remove the `ee` column to
## avoid confusion, and we'll also assign an ID label of `001` and a body mass
## of 87 kg

  body_comp <- info[1, ]
  body_comp$ee <- NULL
  body_comp$wt_kg <- 87

  
## Next, let's determine a Schofield prediction of basal metabolism to impute
## during periods of sleep/non-wear

  body_comp$schofield <- PAutilities::get_ree(
    method = "schofield_wt_ht", sex = "M", age_yr = 55,
    wt_kg = 87, ht_m = 1.85, output = "kcal_day"
  )


## It's time to merge the body composition and energy expenditure data

  ei_data <- merge(body_comp, ee_data)
  

## Next, we want to take our EE estimates (kcal/kg for awake wear periods) and
## convert them to total EE (i.e., no longer relative to body weight, and
## imputing basal metabolic expenditure for periods of sleep/non-wear).
  
## To do this, we'll use some functions from the `dplyr` package. Note that you
## can interpret the `%>%` operator as meaning 'next do ____' or 'then do ___'.
  
## The `%<>%` operator is for assignment (i.e., saves output to the `ei_data` object).
  
  suppressPackageStartupMessages(library(dplyr))
  
  ei_data %<>%
    mutate(across(
      total_hildebrand_linear:total_staudenmayer_rf,
      ~ .x * wt_kg + schofield * ((total_is_NonWear + total_is_Sleep) / 1440),
      .names = "{gsub(\"^total\", \"EE\", .col)}"
    )) %>%
    select(
      any_of(names(body_comp)),
      !c(Date, total_hildebrand_linear:total_staudenmayer_rf)
    ) %>%
    group_by(across(any_of(names(body_comp)))) %>%
    summarise(across(
      everything(), mean, .names = "{gsub(\"^total\", \"mean\", .col)}"
    ), .groups = "drop")

  
  knitr::kable(ei_data)
  ## Note that the `n_days` column refers to the number of days between
  ## body composition measurements, not necessarily the number of days of
  ## accelerometer data included in the analysis (n = 2 in this case)
  

```

#### Phase 3. Calculate EI

At this point, we are ready for the rest of our calculations.

```{r, IB}
## Generate the result
ei_result <-
  ei_data %>%
  mutate(across(
    EE_hildebrand_linear:EE_staudenmayer_rf,
    ~ IntakeBalance::IntakeBalance(
      fm_start, fm_end, ffm_start, ffm_end, ee_per_day = .x
    ),
    .names = "{gsub(\"^EE\", \"EI\", .col)}"
  ))

## Show the result
knitr::kable(ei_result)
```

Voila! The data frame now has new columns that specify EI estimates
for each of the accelerometer-based methods.


## Conclusion

Hopefully this introduction to the `IntakeBalance` package makes it less
intimidating and gets you on your way with implementing it in your own research.
Although there is a lot of code in the above examples (reflecting several steps
you need to implement on your own), it is substantially less than the number of
operations you would have to perform by hand in order to achieve the same
outcome. With a little tweaking for your own projects, our hope is that this
could result in stronger analyses, better quality control, and eventually time
saved as you grow more familiar with using this tool. As noted previously,
please
[let us know if we can improve this vignette](https://github.com/paulhibbing/IntakeBalance/issues/new/choose)
or anything else about `IntakeBalance` to make it more inviting and accessible.
We believe that tools like these have a strong potential to advance science in
the coming years, so we want to provide as much support as we can for increasing
their uptake. Thanks for your interest, and we wish you the best with using
`IntakeBalance`!
